---
title: "Getting started with neuralsbi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with neuralsbi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`neuralsbi` performs **Neural Posterior Estimation (NPE)**: given a prior and a
simulator, it trains a neural network to approximate the Bayesian posterior
`p(theta | x)` — no likelihood required. This vignette walks through the core
workflow and shows how to check that the result is trustworthy.

## The three ingredients

Every SBI problem needs (1) a **prior**, (2) a **simulator**, and (3) an
**observation**. Here is a small linear-Gaussian model whose posterior we happen
to know in closed form, so we can check our answer.

```{r}
library(neuralsbi)

# (1) prior over two parameters
prior <- prior_normal(mean = c(0, 0), sd = 1)

# (2) simulator: an (n x 2) matrix of parameters -> an (n x 2) matrix of data
sigma <- 0.5
simulator <- function(theta) {
  theta + matrix(rnorm(length(theta), sd = sigma), nrow = nrow(theta))
}

# (3) the observation we want to explain
x_obs <- c(1.0, -0.5)
```

## Train an amortized posterior

`npe()` draws parameters from the prior, runs the simulator, and trains a
conditional density estimator. The default is a **Mixture Density Network**
(needs the `torch` back end).

```{r}
fit <- npe(prior, simulator, n_simulations = 5000)
fit
```

The result is *amortized*: it approximates the posterior for **any** observation,
so you only train once.

```{r}
post  <- posterior(fit, x_obs = x_obs)
draws <- sample(post, 10000)

colMeans(draws)          # posterior mean
pairplot(draws)          # joint + marginal view
map_estimate(post)       # MAP point estimate
```

## Did it work? Check against the truth

For this model the posterior is a known Gaussian. Let us compare.

```{r}
d <- 2
Sigma <- solve(diag(d) + diag(d) / sigma^2)
mu    <- as.numeric(Sigma %*% (x_obs / sigma^2))

rbind(analytic = mu, estimated = colMeans(draws))

# classifier two-sample test: ~0.5 => our samples look like analytic samples
z <- matrix(rnorm(10000 * d), ncol = d)
analytic_draws <- sweep(z %*% chol(Sigma), 2, mu, `+`)
c2st(draws, analytic_draws)$accuracy
```

## Calibration when you *don't* know the truth

Usually there is no analytic posterior. **Simulation-based calibration (SBC)**
still tells you whether the posterior is well calibrated: it should produce
uniform rank statistics.

```{r}
res <- sbc(fit, simulator, n_sbc = 300, n_posterior_samples = 500)
plot_sbc(res)                 # flat histogram = calibrated
expected_coverage(res)        # nominal vs empirical credible-interval coverage
```

## A torch-free baseline

If you cannot install `torch`, or want a fast sanity check, use the closed-form
conditional-Gaussian estimator. It is *exact* for linear-Gaussian models.

```{r}
fit_lin <- npe(prior, simulator, n_simulations = 5000,
               density_estimator = "linear_gaussian")
```

## Non-Gaussian posteriors

The MDN is not limited to Gaussian posteriors. See
`system.file("examples", "two_moons.R", package = "neuralsbi")` for the classic
**two-moons** task, whose posterior is bimodal and crescent-shaped, and which the
neural estimator recovers.

## Where to go next

- `?npe`, `?posterior`, `?sbc` for full argument documentation.
- `docs/implementation-plan.md` for the design.
- `docs/verification-roadmap.md` for how methods are validated against Python
  `sbi`.
